---
title: "R for Data Science, 2nd Edition - Chapter 11 Notes"
freeze: true
format:
    html:
        toc: true
        toc-title: Contents
editor_options: 
  chunk_output_type: console
---



# 11 Communication

Here we go from exploratory to expository graphs.

They recommend `The Truthful Art` by Albert Cairo as a guide for data visualization.

```{r}

library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)

```



## 11.2 Labels

Labels help to make a chart expository.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )

```

## 11.3 Annotations

Sometimes it's helpful to label individual observations or groups of observations. You can use `geom_text()`, which functions like `geom_point()`, but it has an additional aesthetic: `label`, which is for textual labels to add to your plot.

```{r}

label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)

label_info
#> # A tibble: 3 × 4
#> # Groups:   drv [3]
#>   displ   hwy drv   drive_type       
#>   <dbl> <int> <chr> <chr>            
#> 1   6.5    17 4     4-wheel drive    
#> 2   5.3    25 f     front-wheel drive
#> 3   7      24 r     rear-wheel drive

```



```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")
#> `geom_smooth()` using method = 'loess' and formula = 'y ~ x'

```


To prevent labels from overlapping with one another and/or with the data, use `geom_label_repel()` from the `ggrepel` package.


```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_y = 2
  ) +
  theme(legend.position = "none")
#> `geom_smooth()` using method = 'loess' and formula = 'y ~ x'

```

Another handy use case for `geom_text_repel()` is to highlight outliers.

```{r}

potential_outliers <- mpg |>
  filter(hwy > 40 | (hwy > 20 & displ > 5))
  
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers, aes(label = model)) +
  geom_point(data = potential_outliers, color = "red") +
  geom_point(
    data = potential_outliers,
    color = "red", size = 3, shape = "circle open"
  )

```

More annotation ideas:

Use `geom_hline()` and `geom_vline()` to add reference lines. We often make them thick (`linewidth = 2`) and white (`color = white`), and draw them underneath the primary data layer. That makes them easy to see, without drawing attention away from the data.

Use `geom_rect()` to draw a rectangle around points of interest. The boundaries of the rectangle are defined by aesthetics `xmin`, `xmax`, `ymin`, `ymax.` Alternatively, look into the `ggforce` package, specifically `geom_mark_hull()`, which allows you to annotate subsets of points with hulls.

Use `geom_segment()` with the arrow argument to draw attention to a point with an arrow. Use aesthetics `x` and `y` to define the starting location, and `xend` and `yend` to define the end location.

Use `annotate()` to highlight just one or two points or ideas.

```{r}

trend_text <- "Larger engine sizes tend to have lower fuel economy." |>
  str_wrap(width = 30)
trend_text
#> [1] "Larger engine sizes tend to\nhave lower fuel economy."

```

Note the annotaiton is accomplised with a `label` geom and a `segment` geom, which are both called within the `annotate` function, in two separate calls to the function.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed")
  )

```


## 11.4 Scales

Control how the aesthetic mappings manifest visually.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

```

Behind the scenes, this is what happens:

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_color_discrete() # chooses colors for each of the class of cars

```

You may want to access these `scale` functions to change things like breaks on the axes, or the key labels on the legend.

You might want to replace the scale altogether.

**Axis ticks and legend keys**.

Collectively, axes and legends are called **guides**. Two arguments affect the appearance of the ticks on the axes and keys on the legend: `breaks` and `labels`.

```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) 

```

`labels` can be set to `NULL`, or a vector of length `breaks`, or in the case of a categorical variables, a named list with mappings to existing level names.

```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  scale_color_discrete(labels = c("4" = "4-wheel", "f" = "front", "r" = "rear"))

```

The `labels` argument can be combined with the `scales` package. Note that the `breaks` argument is on the original scale.

```{r}

# Left
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(labels = label_dollar())

# Right
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, suffix = "K"), 
    breaks = seq(1000, 19000, by = 6000)
  )

```


`label_percent()` is handy when visualizing proportions.


```{r}

ggplot(diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percentage", labels = label_percent())

```


Custom `breaks` arguments can be useful when there isn't much data and you want to highlight specific data points.


```{r}

presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(name = NULL, breaks = presidential$start, date_labels = "'%y")

```



**Legend layout**.

Use the `theme()` setting `legend_position()`. Also use `guides`.

```{r}

base <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

base + theme(legend.position = "right") # the default
base + theme(legend.position = "left")
base + 
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 3))
base + 
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3))

```



```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, override.aes = list(size = 4))) # aes override
#> `geom_smooth()` using method = 'loess' and formula = 'y ~ x'

```


**Replacing a scale**. 

```{r}

# Left
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d()
#> `stat_bin2d()` using `bins = 30`. Pick better value `binwidth`.

# Right
ggplot(diamonds, aes(x = log10(carat), y = log10(price))) +
  geom_bin2d()
#> `stat_bin2d()` using `bins = 30`. Pick better value `binwidth`.

```


`scale_color_brewer()` allows access to palettes that can be seen by people with color blindness.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  scale_color_brewer(palette = "Set1")

```


Redundant shape mapping also improves accessibility

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_color_brewer(palette = "Set1")

```


Specify mapping between colors and variables using `scale_color_manual()`.

```{r}

presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))

```

Viridis color scales are available, and also accessible for continuous, discrete, and binned palettes.

```{r}

df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  labs(title = "Default, continuous", x = NULL, y = NULL)

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_c() +
  labs(title = "Viridis, continuous", x = NULL, y = NULL)

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_b() +
  labs(title = "Viridis, binned", x = NULL, y = NULL)

```

You can zoom in on a portion of the plot using `coord_cartesian()`.

```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  scale_x_continuous(limits = c(5, 6)) +
  scale_y_continuous(limits = c(10, 25))

# Right
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 6), ylim = c(10, 25))

```

```{r}

suv <- mpg |> filter(class == "suv")
compact <- mpg |> filter(class == "compact")

# Left
ggplot(suv, aes(x = displ, y = hwy, color = drv)) +
  geom_point()

# Right
ggplot(compact, aes(x = displ, y = hwy, color = drv)) +
  geom_point()

```


Enforcing common scale limits across plots to enhance comparison.

```{r}

x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_color_discrete(limits = unique(mpg$drv))

# Left
ggplot(suv, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale

# Right
ggplot(compact, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale

```



## 11.5 Themes

Customize non-data elements with a theme.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()

```

Create your own themes, use those included in `ggplot2` or use extension packages like `ggthemes`.

## 11.6 Layout

The `+` operator in the `patchwork` package places two plots next to each other.

```{r}

p1 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) + 
  geom_boxplot() + 
  labs(title = "Plot 2")
p1 + p2

```


More complex layouts with `|` and `/`. 

```{r}

p3 <- ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 3")
(p1 | p3) / p2

```



```{r}

p1 <- ggplot(mpg, aes(x = drv, y = cty, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 1")

p2 <- ggplot(mpg, aes(x = drv, y = hwy, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 2")

p3 <- ggplot(mpg, aes(x = cty, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 3")

p4 <- ggplot(mpg, aes(x = hwy, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 4")

p5 <- ggplot(mpg, aes(x = cty, y = hwy, color = drv)) + 
  geom_point(show.legend = FALSE) + 
  facet_wrap(~drv) +
  labs(title = "Plot 5")

(guide_area() / (p1 + p2) / (p3 + p4) / p5) +
  plot_annotation(
    title = "City and highway mileage for cars with different drive trains",
    caption = "Source: https://fueleconomy.gov."
  ) +
  plot_layout(
    guides = "collect",
    heights = c(1, 3, 2, 4)
    ) &
  theme(legend.position = "top")

```



