---
title: "R for Data Science, 2nd Edition - Chapter 5 Notes"
freeze: true
format:
    html:
        toc: true
editor_options: 
  chunk_output_type: console
---

# 5 Data tidying

```{r}

library(tidyverse)

```

## 5.3 Lengthening data

Use `pivot_longer()`.

```{r}

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  7 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk7      99
#>  8 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk8      NA
#>  9 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk9      NA
#> 10 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk10     NA
#> # ℹ 24,082 more rows

```

- `cols` specifies which columns need to be pivoted, i.e. which columns aren’t variables. This argument uses the same syntax as `select()` so here we could use `!c(artist, track, date.entered)` or `starts_with("wk")`.
- `names_to` names the variable stored in the column names, we named that variable `week.`
- `values_to` names the variable stored in the cell values, we named that variable `rank.`

So `pivot_longer()` starts with identifying which columns are variables, and which are not. For those columns that are not variables, we create a single variable capturing the names of these non-variable columns, and another variable capturing the values in the cells of those non-variable columns, which we must also name.

Some `NA` values may be generated by the structure of the data set. This can be avoided with `values_drop_na = TRUE`.


```{r}

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows

```


The values for `week` are characters. This is a good opportunity to use `parse_number()`.


```{r}

billboard_longer <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(
    week = parse_number(week)
  )
billboard_longer
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered  week  rank
#>   <chr>  <chr>                   <date>       <dbl> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26       1    87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26       2    82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26       3    72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26       4    77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26       5    87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26       6    94
#> # ℹ 5,301 more rows

```


Examining `pivot_longer()`


```{r}

df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
#> # A tibble: 6 × 3
#>   id    measurement value
#>   <chr> <chr>       <dbl>
#> 1 A     bp1           100
#> 2 A     bp2           120
#> 3 B     bp1           140
#> 4 B     bp2           115
#> 5 C     bp1           120
#> 6 C     bp2           125

```

What if column names contain multiple pieces of information?

```{r}

who2
#> # A tibble: 7,240 × 58
#>   country      year sp_m_014 sp_m_1524 sp_m_2534 sp_m_3544 sp_m_4554
#>   <chr>       <dbl>    <dbl>     <dbl>     <dbl>     <dbl>     <dbl>
#> 1 Afghanistan  1980       NA        NA        NA        NA        NA
#> 2 Afghanistan  1981       NA        NA        NA        NA        NA
#> 3 Afghanistan  1982       NA        NA        NA        NA        NA
#> 4 Afghanistan  1983       NA        NA        NA        NA        NA
#> 5 Afghanistan  1984       NA        NA        NA        NA        NA
#> 6 Afghanistan  1985       NA        NA        NA        NA        NA
#> # ℹ 7,234 more rows
#> # ℹ 51 more variables: sp_m_5564 <dbl>, sp_m_65 <dbl>, sp_f_014 <dbl>, …

```

So in this case we have six pieces of information recorded in who2: the country and the year (already columns); the method of diagnosis, the gender category, and the age range category (contained in the other column names); and the count of patients in that category (cell values). 


```{r}

who2 |> 
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"), 
    names_sep = "_",
    values_to = "count"
  )
#> # A tibble: 405,440 × 6
#>   country      year diagnosis gender age   count
#>   <chr>       <dbl> <chr>     <chr>  <chr> <dbl>
#> 1 Afghanistan  1980 sp        m      014      NA
#> 2 Afghanistan  1980 sp        m      1524     NA
#> 3 Afghanistan  1980 sp        m      2534     NA
#> 4 Afghanistan  1980 sp        m      3544     NA
#> 5 Afghanistan  1980 sp        m      4554     NA
#> 6 Afghanistan  1980 sp        m      5564     NA
#> # ℹ 405,434 more rows

```

What if the column names include a mix of of variable values and variable names?

```{r}

household

```

Use the `.value` sentinel value to indicate the first component of the pivoted column name as a variable name in the output. Note the lack of a `values_to` argument here. When you use `.value` in `names_to`, the column names in the input contribute to both values and variable names in the output.

```{r}

household |> 
  pivot_longer(
    cols = !family, 
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )
#> # A tibble: 9 × 4
#>   family child  dob        name 
#>    <int> <chr>  <date>     <chr>
#> 1      1 child1 1998-11-26 Susan
#> 2      1 child2 2000-01-29 Jose 
#> 3      2 child1 1996-06-22 Mark 
#> 4      3 child1 2002-07-11 Sam  
#> 5      3 child2 2004-04-05 Seth 
#> 6      4 child1 2004-10-10 Craig
#> # ℹ 3 more rows

```



## 5.4 Widening data

Use `pivot_wider()` to increase columns and reduce rows.

```{r}

cms_patient_experience |> 
  distinct(measure_cd, measure_title)
#> # A tibble: 6 × 2
#>   measure_cd   measure_title                                                 
#>   <chr>        <chr>                                                         
#> 1 CAHPS_GRP_1  CAHPS for MIPS SSM: Getting Timely Care, Appointments, and In…
#> 2 CAHPS_GRP_2  CAHPS for MIPS SSM: How Well Providers Communicate            
#> 3 CAHPS_GRP_3  CAHPS for MIPS SSM: Patient's Rating of Provider              
#> 4 CAHPS_GRP_5  CAHPS for MIPS SSM: Health Promotion and Education            
#> 5 CAHPS_GRP_8  CAHPS for MIPS SSM: Courteous and Helpful Office Staff        
#> 6 CAHPS_GRP_12 CAHPS for MIPS SSM: Stewardship of Patient Resources

```


`pivot_wider()` has the opposite interface to `pivot_longer()`: instead of choosing new column names, we need to provide the existing columns that define the values (`values_from`) and the column name (`names_from`):


```{r}

cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
#> # A tibble: 500 × 9
#>   org_pac_id org_nm                   measure_title   CAHPS_GRP_1 CAHPS_GRP_2
#>   <chr>      <chr>                    <chr>                 <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          63          NA
#> 2 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          NA          87
#> 3 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          NA          NA
#> 4 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          NA          NA
#> 5 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          NA          NA
#> 6 0446157747 USC CARE MEDICAL GROUP … CAHPS for MIPS…          NA          NA
#> # ℹ 494 more rows
#> # ℹ 4 more variables: CAHPS_GRP_3 <dbl>, CAHPS_GRP_5 <dbl>, …

```

If you still have multiple rows per unit of interest, organization in this case, then the `pivot_wider()` didn't work.

The problem is we need to tell `pivot_wider()` which column or columns have values that uniquely identify each row, in this case those are the variables starting with `"org"`:

```{r}

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
#> # A tibble: 95 × 8
#>   org_pac_id org_nm           CAHPS_GRP_1 CAHPS_GRP_2 CAHPS_GRP_3 CAHPS_GRP_5
#>   <chr>      <chr>                  <dbl>       <dbl>       <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICA…          63          87          86          57
#> 2 0446162697 ASSOCIATION OF …          59          85          83          63
#> 3 0547164295 BEAVER MEDICAL …          49          NA          75          44
#> 4 0749333730 CAPE PHYSICIANS…          67          84          85          65
#> 5 0840104360 ALLIANCE PHYSIC…          66          87          87          64
#> 6 0840109864 REX HOSPITAL INC          73          87          84          67
#> # ℹ 89 more rows
#> # ℹ 2 more variables: CAHPS_GRP_8 <dbl>, CAHPS_GRP_12 <dbl>

```


How does `pivot_wider()` work?


```{r}

df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

```

Take the values from the `value` column and the column names from the `measurement` column.

```{r}

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
#> # A tibble: 2 × 4
#>   id      bp1   bp2   bp3
#>   <chr> <dbl> <dbl> <dbl>
#> 1 A       100   120   105
#> 2 B       140   115    NA

```

Any column that does not go into the `names_from` or `values_from` argument determines the number of rows. These are called the `id_cols`.

What if we have duplicate values in the column giving the new column names and in the column giving the new value names? We get list-columns.

```{r}

df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "A",        "bp1",    102,
  "A",        "bp2",    120,
  "B",        "bp1",    140, 
  "B",        "bp2",    115
)


df |>
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
#> Warning: Values from `value` are not uniquely identified; output will contain
#> list-cols.
#> • Use `values_fn = list` to suppress this warning.
#> • Use `values_fn = {summary_fun}` to summarise duplicates.
#> • Use the following dplyr code to identify duplicates.
#>   {data} |>
#>   dplyr::summarise(n = dplyr::n(), .by = c(id, measurement)) |>
#>   dplyr::filter(n > 1L)
#> # A tibble: 2 × 3
#>   id    bp1       bp2      
#>   <chr> <list>    <list>   
#> 1 A     <dbl [2]> <dbl [1]>
#> 2 B     <dbl [1]> <dbl [1]>

```

Follow the hint in the warning to locate the duplicate:



```{r}

df |> 
  group_by(id, measurement) |> 
  summarize(n = n(), .groups = "drop") |> 
  filter(n > 1)
#> # A tibble: 1 × 3
#>   id    measurement     n
#>   <chr> <chr>       <int>
#> 1 A     bp1             2

```

```{r}

df

```


























