---
title: "R for Data Science, 2nd Edition - Chapter 9 Notes"
freeze: true
format:
    html:
        toc: true
editor_options: 
  chunk_output_type: console
---

# 9 Layers

```{r}

library(tidyverse)

```

## 9.2 Aesthetic mappings

```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()

# Right
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
#> Warning: The shape palette can deal with a maximum of 6 discrete values because more
#> than 6 becomes difficult to discriminate
#> ℹ you have requested 7 values. Consider specifying shapes manually if you
#>   need that many of them.
#> Warning: Removed 62 rows containing missing values or values outside the scale range
#> (`geom_point()`).

```


```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()
#> Warning: Using size for a discrete variable is not advised.

# Right
ggplot(mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point()
#> Warning: Using alpha for a discrete variable is not advised.

```

Arguments outside of `aes()` do not map onto variables to determine appearance.


```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")

```


## 9.3 Geometric objects


Same data, different geoms:

```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()

# Right
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_smooth()
#> `geom_smooth()` using method = 'loess' and formula = 'y ~ x'

```

Not every aesthetic works with every geom. You can't set the `shape` or a line geom.

```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_smooth()

# Right
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) + 
  geom_smooth()

```

Instead of giving a shape to a line geom, ggplot split the lines by `drv`.

For geoms that produce a single geometric object, you can use the `group` aesthetic to draw multiple objects. Note that the `group` aesthetic does not add a legend or distinguishing features to the geoms.


```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(aes(linetype = drv))

```

Placing mappings in a geom function overwrites the global mappings for that layer only. This means you can have different aesthetics in different layers.



```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = class)) + 
  geom_smooth()

```

You can use this idea to specify different data for each layer.


```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_point(
    data = mpg |> filter(class == "2seater"), 
    color = "red"
  ) +
  geom_point(
    data = mpg |> filter(class == "2seater"), 
    shape = "circle open", size = 3, color = "red"
  )

```



```{r}

# Left
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)

# Middle
ggplot(mpg, aes(x = hwy)) +
  geom_density()

# Right
ggplot(mpg, aes(x = hwy)) +
  geom_boxplot()

```



## 9.4 Facets


`facet_wrap()` splits a plot into subplots that each display one subset of the data based on a categorical variable.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_wrap(~cyl)

```


To facet with a combination of two variables, switch to `facet_grid()`.


```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl)

```


Let the x and y axes vary across plots with `scales = "free"`.


```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl, scales = "free")

```

Use the `.` to facet horizontally?

```{r}

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)


```



```{r}

ggplot(mpg, aes(x = displ)) + 
  geom_histogram() + 
  facet_grid(drv ~ .)

ggplot(mpg, aes(x = displ)) + 
  geom_histogram() + 
  facet_wrap(~drv)

ggplot(mpg, aes(x = displ)) + 
  geom_histogram() +
  facet_grid(. ~ drv)



```



## 9.5 Statistical transformations

When new values have to be calculated for the plot.

-   Bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.

-   Smoothers fit a model to your data and then plot predictions from the model.

-   Boxplots compute the five-number summary of the distribution and then display that summary as a specially formatted box.

```{r}

ggplot(diamonds, aes(x = cut)) + 
  geom_bar()

```

The algorithm used to calculate new values for a graph is called a **stat**, short for statistical transformation.

You can see which stat a geom uses by looking at default arguments, and the documentation for a given stat can be found on the geom help page. See how `stat_count()` is documented on the `geom_bar()` help page.

Every geom has a default stat, and every stat has a default geom.

You might need to use the stat explicitly to:

1. override the default stat


```{r}

diamonds |>
  count(cut) |>
  ggplot(aes(x = cut, y = n)) +
  geom_bar(stat = "identity")

```

2. Override default mapping from transformed variables to aesthetics.

For example, a bar chart of proportions instead of counts is possible like this:


```{r}

ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) + 
  geom_bar()

```


3. You might want to draw greater attention to the statistical transformation in your code. Here is an example with `stat_summary()`, which summarizes the `y` value for each unique `x` value plotted.

```{r}

ggplot(diamonds) + 
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )

```



## 9.6 Position adjustments

```{r}


```




```{r}



```



```{r}



```




```{r}



```



```{r}



```



```{r}



```



```{r}



```




```{r}



```



```{r}



```

